\chapter*{Acknowledgements}
\phantomsection
\addcontentsline{toc}{chapter}{Acknowledgements}
\vspace*{\fill}
\noindent This thesis would not have been possible without the support, guidance, and encouragement of many people.

First, I would like to express my deepest gratitude to my primary supervisor, Prof. Sunil Gupta. His expertise, patience, and thoughtful advice have been essential during this journey. I am also sincerely thankful to my co-supervisor, Dr. Hung-The Tran, for his valuable insights, constructive feedback, and constant support, which have greatly improved my work.

I am especially grateful to Dr. Alistair Shilton for his important contributions to my physics informed black-box optimization project. His guidance and suggestions have helped me refine my ideas and approach. I also appreciate the time he dedicated to carefully reviewing this thesis.

I want to thank my friends and colleagues at the Applied Artificial Intelligence Institute (A\textsuperscript{2}I\textsuperscript{2}) for creating such a welcoming and supportive environment. Their encouragement and collaboration have made this experience both rewarding and enjoyable.

Lastly, I am deeply thankful to my family—my parents, my brother, and my partner. Their love, belief in me, and constant support have been my greatest source of strength. They have stood by me through every challenge and celebrated every achievement, and this thesis would not have been completed without them.

To everyone who has supported me along the way, I extend my heartfelt thanks. Your kindness and encouragement mean more to me than words can express.
\vspace*{\fill}
\cleardoublepage
\pagestyle{plain}